---
title: "p8105_mtp_xy2395"
author: "Jack Yan"
date: "10/16/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(patchwork)
```

```{r, eval=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

```

# Data Cleaning and EDA

The cleaning step is described below in the code chunk:

```{r}
# Import and clean the dataset
motion_data = 
  read_csv("./data/p8105_mtp_data.csv") %>% 
  janitor::clean_names() %>% 
  # all the original activity values are preserved in a single variable
  gather(key = minute, value = activity, activity_1:activity_1440) %>% 
  mutate(
    # truncate the extra 'activity_' string in the `minute` variable and change it to integer
    minute = str_replace(minute, "activity_", "") %>% as.integer(),
    # change the 'day' variable to factor and relevel them
    day = as.factor(day),
    day = forcats::fct_relevel(day, c("Sunday", "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
  )
```

To facilitate further analysis (grouping), I created two new variables, `day_num` and `hour` (described below). 

```{r}
motion_data = 
  motion_data %>% 
  mutate(
    # `day_num` variable denotes number of days past the beginning of the study
    day_num = ((week - 1) * 7 + as.numeric(day)) %>% as.integer(),
    # `hour` variable denotes 24 hours in a day
    hour = as.integer((minute - 1) / 60) %>% as.factor()
  ) %>% 
  # The variables and observations are arranged in a reasonable order.
  select(day_num, week, day, minute, hour, activity) %>% 
  arrange(week, day, minute) 
```

The final dataset has `r nrow(motion_data)` observations and `r ncol(motion_data)` variables. The `activity` variable contains all the activity counts in each minute. Variables `day`, `minute`, `hour`, `day_num` and `hour` colletively specify the time when each observation was made.

```{r }
# A summary of the variables in the cleaned dataset 
str(motion_data)
```

## EDA

A visualization of distribution of daily activity was made and shown in the next section. Interestingly, for some given days, activity measurement was close to zero, incidating no activity recorded for the person that day.

# Daily Total Activity 

## Daily Trend

As shown in the first plot, daily total activity increased through the first and last 1/3 parts of the period, and witness a slight decline in the middle. The weekly aggregate activity followed a similar pattern. Generally, we can indicate an increased daily activity over time.

```{r}
# calculate daily total activity 
total_activity = 
  motion_data %>% 
  group_by(day_num) %>% 
  # total activity variable: `sum`
  summarize(sum = sum(activity))

# plot trend across days
total_activity %>% 
  # To see if this participant became more active over time, 
  # plot the daily total activity among days.
  ggplot(aes(x = day_num, y = sum)) +
    geom_point() +
    # The trend lines are plotted using `geom_smooth` with `loess` method
    geom_smooth(method = 'auto', se = FALSE) +
    labs(
      title = "Distribution and Trend of Daily Activity",
      x = "Day #",
      y = "Daily Total Activity"
    )

```

## Daily Total Activity by Day of the Week 

The distribution of activity differs substantially among different days in a week. For Sunday, Saturday and Friday, a larger portion of the area below the density curve shifts to a higher amount, indicating more activity on those days. Also, the peak for Sunday is lower than other days, which indicates a higher diversity. Therefore, daily activity differ by day in a week.

```{r fig.width = 11, fig.height = 6}
# density distribution of daily total activity by day of the week
total_activity_dow = 
  motion_data %>% 
  group_by(week, day, day_num) %>%
  summarize(sum = sum(activity)) %>% 
  ggplot(aes(x = sum, y = day)) + 
  geom_density_ridges(scale = .85) +
  labs(
    y = "Day of the Week",
    x = "Activity",
    title = "Distribution of Daily Activity by Day of the Week"
  ) +
   viridis::scale_color_viridis(
    name = "Day of the Week", 
    discrete = TRUE
  )

# Plot trend of daily total activity grouped by day of the week
# See if the progress of time affects daily activity 
trend_days_by_dow = 
  motion_data %>% 
  group_by(week, day, day_num) %>%
  summarize(sum = sum(activity)) %>% 
  ggplot(aes(x = week, y = sum, color = day)) +
    geom_point(alpha = 0.4) +
   # show the trend of daily activity using `geom_smooth` with the `lm` method
    geom_smooth(method = 'lm', se = FALSE) +
    labs(
      title = "Trend of Daily Activity by Day of the Week",
      x = "Week #",
      y = "Total Activity"
    ) +
   viridis::scale_color_viridis(
      name = "Day of the Week", 
      discrete = TRUE
   )

# combine and show the two plots
total_activity_dow + trend_days_by_dow
```

The second plot shows the trend of activity grouped by day of the week. The daily activity increased for all days over time, but their slopes differ. Activity increased most on Sundays.

# Activity over the course of a day

## 24-hour activity profiles

In this section, we start to investigate how activity counts vary in 24 hours each day.    

```{r}
  # create a density ridges plot to see the distribution of hourly activity
  motion_data %>% 
  # group the dataset by `hour` and `day_num`
  group_by(day_num, hour) %>% 
  summarize(hour_activity = sum(activity)) %>% 
  # values > 40000 in `hour_activity` was filtered out,
  # because some outliers in the `hour_activity` variable overly extend the X scale and therefore affect readability
  filter(hour_activity <= 40000) %>% 
  ggplot(aes(y = hour, x = hour_activity)) +
    # draw distribution of hourly activity in different days
    geom_density_ridges() +
    labs(
      x = "Hour Total Activity",
      y = "Hour",
      title = "Density Distribution of 24-hour activity"
    )

```

We can see from this plot what time everyday this person tended to be more active and even indicate the person's daily routine. For example, from 10 pm to 7 am, the person tended to be inactive.

## Effects of time and day of the week on 24-hour activity profiles

As shown in the first plot, the person tends to be most active around 12 pm on Sundays, and is also very active around 6 pm on Fridays and around 12 pm on Saturdays. Additionally, on Thurdays, Mondays and Fridays, the peak of activity is round 6 pm, which is different from other days.

```{r fig.width = 11, fig.height = 6}
# visualize effects of time and day of the week on 24-hour activity profiles
# effect of day in the week on 24-hour activity profiles
activity_profile_dow = 
  motion_data %>% 
  # group the dataset by day of the week and hour
  group_by(day, hour) %>%
  # calculate the _average hourly activity_ by summing up the activity counts in each group and divide it by the total number of weeks
  summarize(avg_activity = sum(activity) / 47) %>% 
  ggplot(aes(x = hour, y = avg_activity, group = day, color = day)) +
    # The 24-hour trend line for activity was plotted using the default "loess" method. 
    geom_smooth(se = FALSE) +
    labs(
      x = "Hour",
      y = "Hourly Activity",
      title = "24-hour activity profile by day of the week"
    ) +
    viridis::scale_color_viridis(
    name = "Day of the Week", 
    discrete = TRUE
  ) +
  theme(legend.position = "bottom")

# effect of week number on 24-hour activity profiles
activity_profile_week = 
  motion_data %>% 
  group_by(week, day_num, hour) %>%
  summarize(sum_activity = sum(activity)) %>% 
  ggplot(aes(x = hour, y = sum_activity, group = week, color = week)) +
    geom_smooth(se = FALSE, size = 0.5) +
    labs(
      x = "Hour",
      y = "Total Activity",
      title = "24-hour activity profile across time"
    ) +
    viridis::scale_color_viridis(
    name = "Week", 
    discrete = FALSE
    ) +
    theme(legend.position = "bottom")

activity_profile_dow + activity_profile_week
```

To examine the effect of time on 24-hour activity profiles, I used average activity data across weeks. The trend lines are grouped by weeks, each showing the average 24-hour profile in a single week. 

We can see that with the progress of time, activity tends to have a higher peak, indicating more activity.

